---
title: 'CANSSI/UBC Datathon: simulated global fish catch'
author: "Isabella Morgante"
output:
  html_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data availability statement


The dataset for this activity is available in the file *fish_data.csv* shared via Github (https://github.com/MarieAugerMethe/CANSSI_HMM_wk_dthon_2025). It can be used for our CANSSI/UBC datathon and shared for educational purposes with accreditation (Isabella Morgante, Juliano Palacios Abrantes, Muhammed Oyinlola & William Cheung). It cannot be used for any publication without written permission. Please contact Isabella Morgante (i.morgante@oceans.ubc.ca) & Juliano Palacios Abrantes (j.palacios@oceans.ubc.ca) for more information. 

Let's load some of the packages needed.

```{r, warning=FALSE, message=FALSE}
library(tidyverse)
```


## Description of the dataset

This dataset contains simulations of total annual global fish catch and abundance for three fish species from 1850 - 2010: European anchovy, Chub mackerel, and European pilchard. The simulations were run using the Dynamic Bioclimatic Envelope Model (DBEM), a global fisheries model by Cheung et al. (2011) and fishing effort data from the Fisheries Model Intercomparison Project (FishMIP).

For each species, 2 simulations are available – one where there is no fishing occurs and a second where fishing is applied. All simulations were run under the same environmental conditions. The annual average fishing mortality (a proportion of maximum sustainable yield)  and the annual average sea surface temperature (°C) each species was exposed to is also provided.


Let's read the data into R and have a quick peak at it.

```{r data.load}
fish_sim <- read.csv("fish_data.csv")
head(fish_sim)
```


As you can see, the dataset contains these variables:

- **Abd:** Total fish abundance (with fishing)
- **Abd_noCatch:** Total fish abundance (with no fishing)
- **Catch:** Total fish catch
- **SST:** mean sea surface temperature (°C)
- **FMort:** mean fishing mortality
- **Year:** year of data
- **TaxonID:** identification code of the fish species
- **TaxonName:** Scientific name of the fish species
- **CommonName:** Common name of the fish species

### Quick visualization

```{r}
# Scale to be able to plot both axes
sc <- 100
ggplot(data = fish_sim, aes(x = Year)) +
  geom_line(aes(y = Abd, colour = CommonName), size = 1) +
  geom_line(aes(y = Catch*sc, colour = CommonName), size = 0.5, lty = 2) +
  scale_y_continuous(
    
    # Features of the first axis
    name = "Fish abundance",
    
    # Add a second axis and specify its features
    sec.axis = sec_axis(~.*sc, name="Fish catch (dashed)")
  )
  
```

## Datathon goal

The goals are (1) to provide a road map of how to tackle the questions listed below, (2) attempt to complete at least the first step of your road map, and (3) provide interpretation of the results. Make a quick 5-minute presentation explaining what your team did.

### Questions:

XX LIST THE QUESTIONS YOU WANT TO ANSWER WITH THE DATA. KEEP IN MIND THAT HIDDEN MARKOV MODEL WILL TAKE A TIME SERIES AND FIND HIDDEN STATES WITHIN IT. USUALLY YOU WILL HAVE 3-4 HIDDEN STATES.

## Acknowledgments

If the analysis you developed with your team,  or a derivative of it, is used in your thesis or in a publication, please add the name of our research team and of your teammates in the acknowledgements. Please also let Marie Auger-Méthé (auger-methe@stat.ubc.ca) and Vianey Leos Barajas (vianey.leosbarajas@utoronto.ca) know, as we will be excited to see that our datathon has provided you with concrete help! The name of our team is: **CANSSI: Advancing Statistical Methods for the Analysis of Complex Biologging Data Collected from Humans and Animals**.


